### 2020-01-03 manuscript/manuscript.rmd
update manuscript

```{r preliminaries}
library( tidyverse )
library( magrittr )
library( viridis )
library( knitr )
library( gridExtra )

source( "functions.R" )

@@

^3^Department of Biological Sceince, Vanderbilt University

@@

## Abstract
Phylogenomic studies based on hundreds to thousand of genes have greatly improved our understanding of evolutionary relationship among major animal lineages. However, a considerable debate still remains regarding the first diverging animal lineage, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life. Here, we systematically explore the mechanisms (evolutionary models, gene sampling and outgroup choice) that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies. We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices. Moreover, We find Porifera-sister can only be recovered in several data matrices using site-heterogeneous CAT model with inclusive of the most closely related outgroups, whereas Ctenophora-sister is recovered in all other cases. We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between models and outgroup choice (in site-heterogeneous CAT models) . Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses. Our results provide a comprehensive overview of current understanding of the first diverging animal branch and why different studies yield different answers. Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$. A detailed description of each component is in Supplementary text.
Briefly, (1) the exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR, as they have been used regularly in phylogenomic analyse; (2) While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled, such as Site homogeneous rates, Gamma-rate heterogeneity; (3) the vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids, such as empirical site-homogeneous, estimated site-homogeneous, CAT site heterogeneous [@Lartillot:2004dq], and C10-C60 models (C models) . (4) data-recoding method recode 20 amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation. Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models.

Models can be assembled by selecting different options for all these different components. The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention. For example, on the questions considered here, Poisson and GTR exchangeability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangeability matrix. CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model.To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma). If there are no modifiers, then it is implied that site homogeneous models are used.

@@

High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes (*e.g.*, [@whelan2017ctenophore; @simion2017large]) to explore the root of metazoan phylogeny. Although the number of sampled genes might be different across studies, it should expect that the sampled genes should be conserved across most animal genomes and most data matrices should share a core set of orthology genes (OGs) (*e.g.*, BUSCO genes). Surprisingly, gene composition and matrix overlap across different studies has been rarely assessed, and most studies relied on their own bioinformatic pipeline for determination of OGs (Supplementary text).

@@

Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of the root position of animal phylogeny (**Fig XXOverview.C**). Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [e.g. @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Opisthokonta) outgroup taxa tend to recover Ctenophora-sister, whereas high support of Porifera-sister often recovered when site-heterogeneous CAT models were used [@pisani2015genomic]. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.

@@

Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the deep animal phylogeny [@philippe2011resolving]. Feuda *et al*. (2017) [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in data matrices. By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent. By reanalyzing Chang et al. 2015 [@chang2015genomic] and Whelan et al. 2015 [@whelan2015error], they report that posterior predictive analyses indicate recoded analyses have better model adequacy than non-recoding analyses, and that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved]. However, the performance of the recoding methods has never been tested empirically against the non-recoding method, and a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six].

@@

Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position that needs to be explored. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting] ). Importantly, everal earlier studies only include less than five ctenophore species. The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of animal root position. Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other data matrices.

@@

Gene sampling plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes for deep animal phylogney. The basic idea is that the concatenation assumes that phylogenetic signal from genes that do not share the species phylogeny will be overwhelmed by the signal from the majority of genes whose genealogy mirrors that of the species evolutionary history [@lanier2012recombination]. However, large phylogenomic data matrices potentially contain genes with conflicting or weak phylogenetic signal that can lead to incongruence phylogenomic results, especially when analyzing deeper node represents ancient divergence in the tree [@salichos2013inferring]. Thus, it is important to evaluate the gene composition and matrix overlap to explore level of congruence of gene sampling across different matrices. It should be noted that early studies relied on EST via Sanger sequencing contains large missing data compared to other ones.

We first used the Metazoa BUSCO dataset to evaluate gene composition from all data matrices and the result is unexpected. The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco]. Moreover, the BUSCO dataset also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics] ). Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices given their nature of conservation and ubiquitous. However, to our surprise, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig. XXBUSCO_annotations).

Previous reanalyses have mainly focused on outgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated. Our initial intent of comparing matrix overlap across matrices was to construct a new matrix with shared taxon and gene sampling to test specific hypotheses about differences in support. The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of OGs. However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig XXAlignment overlap). These results indicated that different studies essentially sampled different regions of the genomes with little overlap.

@@

busco_overrepresented

p1 = ggplot( matrix_summary ) +
	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
	geom_abline(slope=1,intercept=0) +
	coord_fixed()

@@

p2 = ggplot( matrix_summary ) + geom_point( aes(x=matrix, y=n_busco_partitions/n_partitions*100, col=manuscript) ) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))

@@

**Fig XXBUSCO_annotations.** The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions.

@@

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. support_published_analyses and Table perviouly_published_analyses). Among these `r length(unique(analyses_published$manuscript))` studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 supports Ctenophora-sister (Table PAPERS). Moreover, three of them are reanalyses papers without generating any new data matrices. We found that all past phylogenetic analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of the root of animal phylogeny) mianly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig. support_published_analyses).

@@

Here we reanalyzed 17 from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table datasets). We also progressively trimmed each data matrix with different level of outgroup choice if applicable. We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree (except for large matrices including Hejnol2009, Borrowiec_Total1080 and Simion2017 matrices). We found in all cases, C60+ WAG/LG models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree. It should be noted that we didn't include GTR + C models in model testing since it is computational too expensive. The GTR+C models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.
With the exception of Moroz2014 matrix that supports neither Porifera or Ctenophora-sister in every models, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.

@@

Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister. This has been addressed with CAT models. IQtree provides a new family of C models that also address site heterogeneity. Given the extensive computational cost and concerns about overparameterization of CAT models, we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results. This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses. For computational efficiency, we first only applied Poisson+CAT models in every matrix with only Choanoflagellata as outgroup. It should be noted that most Phylobayes runs were converged, several large matrices have not reached convergence after at least a month's computational time.

Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in Choanimalia matrices. Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d matrix. Consistent with previous study, we recovered strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, Nosenko2013 and Whelan2015. In contrast, we found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017. Ctenophora-sister were strongly supported in all other matrices. For the matrices have conflicted results between Poisson-CAT and C models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice will affect phylogenetic inference under CAT model. We found strong support of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups. Thus, our results show that both model choice and outgroup sampling play a major role in incongruent results of the root position of animal phylogeny in several key data matrices.

@@

The distribution of phylogenetic signal over the root position of animal phylogeny has never been quantified with site-heterogeneous models. To further explore the effect of model selection and outgroup choices in the root position of animal phylogeny, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three key data matrices with two outgroup choice and four models in both ML and BI framework (Fig XXPhylogenetic signal). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models. Moreover, the outgroup choice has little impact in general on the distribution of the support phylogenetic signal in site-homogeneous models. This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].

In contrast, the distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in two matrices from Ryan_2013 that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig. phylogenetic_signal B). It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of root position of animal phylogeny. In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT model and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal. We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled genes) when exclusive of more distantly related outgroups with the Poisson-CAT models. The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister increases in Holozoa and Choanimalia matrices.

We also examined the level of congruence of genes that are supported in each hypothesis across different analyses in Whelan2017 matrices. Consistent with the summary of \Delta$GLS analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice). In other words. our results strongly indicated that the distribution of phylogenetic signal of the root of animal phylogeny is largely sensitive to model selection (site-homogeneous vs. site-heterogeneous models). More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C models in these data matrices. Thus, it is clear that recovering Porifera-sister requires more constrains including model, outgroup choice, and even with gene sampling.

```{r phylogenetic_signal}
#au_tests %>%
#  unite(Matrix, c(manuscript, matrix, model)) %>%
#  group_by(Matrix, result) %>%
#  count() %>%
#  group_by(Matrix) %>%
#  mutate(percent = prop.table(n)) %>%

@@

library(ggpubr)
theme_set(theme_pubr())
au_tests = read_tsv("../trees_new/AU_test/summary.tsv")

>>>>

```{r preliminaries}
library( tidyverse )
library( magrittr )
library( viridis )
library( knitr )
library( gridExtra )
library(ggpubr)
theme_set(theme_pubr())
source( "functions.R" )

@@

^3^Department of Biological Science, Vanderbilt University

@@

## Abstract
Phylogenomic studies based on hundreds to thousand of genes have greatly improved our understanding of evolutionary relationship among major animal lineages. However, a considerable debate still remains regarding the first diverging animal lineage, with Ctenophora-sister and Porifera-sister emerging as the two primary hypotheses, comprising one of the most difficult node to resolve in animal tree of life. Here, we systematically explore the mechanisms (evolutionary models, gene sampling and outgroup choice) that potentially cause the incongruent results by synthesizing data and results from all previous phylogenomic analyses, and perform new analyses in these data in a consistent manner to characterize the differences between studies. We find the gene sampling varies across studies, indicating that different regions of the genomes are sampled with little overlap among data matrices. More Importantly, We find Porifera-sister can only be recovered with more constraints, including using site-heterogeneous CAT model with inclusive of the most closely related outgroups in several matrices, whereas Ctenophora-sister is recovered in all other cases. We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences of distribution of phylogenetic signal between models and outgroup choice (especially CAT models). Last, we show that recoding can be a problematic method for addressing compositional variation in phylogenetic analyses. Our results provide a comprehensive overview of current understanding of the first diverging animal branch and why different studies yield different answers. Rather than embracing one hypothesis over another, we hope our efforts provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$. A detailed description of each component is in Supplementary text. Briefly, (1) the exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices have been used in the studies under consideration here include: Poisson (or F81), WAG, LG, GTR, as they have been used regularly in phylogenomic analyse; (2) While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled, such as site homogeneous rates and Gamma-rate heterogeneity; (3) the vector of equilibrium frequencies $\Pi$ describes the stationary frequency of amino acids, such as empirical site-homogeneous, estimated site-homogeneous, CAT site-heterogeneous [@Lartillot:2004dq], and C10-C60 models (C models). (4) data-recoding method recode 20 amino acids are recoded into six groups based on function to account for both compositional heterogeneity and substitution saturation. Several recoding strategies have been proposed, including Dayhoff 6-state recoding, S&R 6-state recoding, KGB 6-state recoding based on exchangeability matrix of models [@feuda2017improved].

Models can be assembled by selecting different options for all these different components. The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention. For example, on the questions considered here, Poisson and GTR exchangeability matrices have only been used in combination with CAT site-heterogeneous models of equilibrium frequency. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangeability matrix. CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma). If there are no modifiers, then it is implied that site homogeneous models are used.

@@

High-throughput sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes to explore the root of metazoan phylogeny (Table 1). Although the number of sampled genes might be different across studies, it is reasonable to expect that a conserved core set of orthology genes (OGs) should be present across most animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]). Surprisingly, gene composition and matrix overlap has been rarely assessed across studies and most studies relied on their own bioinformatic pipeline for determination of OGs (Supplementary text) for animal phylogeny.

@@

Outgroup selection (the taxa used to root the ingroup taxa) can potentially affect the phylogenetic inference of the root position of animal phylogeny (**Fig XXOverview.C**). Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas high support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous CAT models [@pisani2015genomic]. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.

@@

Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the deep animal phylogeny [@philippe2011resolving]. Feuda *et al*. (2017) [@feuda2017improved] has proposed to use data-recoding method to reduce compositional heterogeneity in data matrices. By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent. By reanalyzing Chang et al. 2015 [@chang2015genomic] and Whelan et al. 2015 [@whelan2015error], they report that posterior predictive analyses indicate recoded analyses have better model adequacy than non-recoding analyses, and that Porifera-sister was strongly favored under all recoding strategies using GTR-CAT models [@feuda2017improved]. However, how the recoding methods impact phylogenetic inference has not been well explored. Moreover, a recent simulation study has suggested that non-recoding approaches significantly outperformed recoding strategies [@hernandez2019six].

@@

Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position that needs to be explored. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in bilaterians compared to other major animal lineages (Fig XXTaxon_composition). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting] ). Importantly, several earlier studies only include less than five ctenophore species. The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies a lot. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of animal root position. Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other data matrices (Table1, Fi. XXTaxon_composition).

@@

Gene sampling plays an essential role in phylogenomic analyses, and most often rely on the concatenation of sampled genes for deep animal phylogeny. It has been suggested that the concatenation assumes that phylogenetic signal from genes that do not share the species phylogeny will be overwhelmed by the signal from the majority of genes whose genealogy mirrors that of the species evolutionary history [@lanier2012recombination]. However, large phylogenomic data matrices potentially contain genes with conflicting or weak phylogenetic signal that can lead to incongruence phylogenomic results, especially when analyzing deeper node represents ancient divergence in the tree [@salichos2013inferring]. Thus, it is important to evaluate the gene composition and matrix overlap to explore level of congruence of gene sampling across different matrices. It should be noted that early studies relied on EST via Sanger sequencing contains large missing data compared to other ones (*e.g.*, Dunn2008 and Hejnol2009).

We first used the Metazoa BUSCO dataset to evaluate gene composition from all data matrices and the result is unexpected. The animal BUSCO dataset contains 978 single-copy OGs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco]. Moreover, the BUSCO dataset also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics] ). Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices. However, to our surprise, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Fig. XXBUSCO_annotations).

Previous reanalyses have mainly focused on outgroup selection and/or model fits in different data matrices, and the matrix overlap across different studies has been rarely investigated. Our initial intent of comparing matrix overlap across matrices was to construct a new matrix with shared taxon and gene sampling to test specific hypotheses about differences in support. The hypothesis is that although the bioinformatics pipeline used for OG determination is different, most larger data matrices should share a core set of OGs. However, our results show that surprisingly little overlap across all the data matrices from different studies (Fig XXAlignment overlap). Thus, these results suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices.

@@

p1 = ggplot( matrix_summary ) +
	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
	geom_abline(slope=1,intercept=0) + labs(x ="Number of gene partitions", y = "Number of BUSCO genes")
	coord_fixed()

@@

p2 = ggplot( matrix_summary ) + geom_point( aes(x=matrix, y=n_busco_partitions/n_partitions*100, col=manuscript) ) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1)) + labs(x ="Matrix", y = "Percentage of  BUSCO genes")

figure <- ggarrange(p1, p2,
                    labels = c("A", "B"),
                    ncol = 1, nrow = 2, widths=c(1,1), heights=c(1,1),common.legend = TRUE, legend = "right")

@@

**Fig XXBUSCO_annotations.** BUSCO annotation and representation in each data matrix. (A). The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions. (B). The percentage BUSCO annotations in each matrix.

@@

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. support_published_analyses and Table perviouly_published_analyses). Among these `r length(unique(analyses_published$manuscript))` studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 supports Ctenophora-sister (TablexxPAPERS). Moreover, three of them are studies of reanalysis without generating any new data matrices. We found that all past phylogenetic analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups with the use of site-heterogeneious CAT (Poisson+CAT+G or GTR+CAT+G) models. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of the root of animal phylogeny) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT+G and recoding methods with GTR+CAT+G models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only Choanofalgellates are used as outgroups (Fig. support_published_analyses).

```{r overview of studies}
table_Papers = read_tsv("../data_processed/tables/previously_published_manuscripts.tsv")
	kable(table_Papers, caption= "Table XX. Overview of previous phylogenomic studies related to the root position of animal phylogeny")

@@

Here we reanalyzed 17 from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table datasets). We also progressively trimmed each data matrix with different level of outgroup choice if applicable. We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C models using ModelFinder in IQtree (except only LG+C60+F+G models are used large matrices including Hejnol2009, Borrowiec_Total1080 and Simion2017 matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree. It should be noted that we didn't include GTR + C models in model testing since it is computational too expensive. The GTR+C models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous + C models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG+G, GTR+G and Poisson+C60+F+G. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups.

With the exception of Moroz2014 matrix that supports neither Porifera or Ctenophora-sister in every models, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups.

@@

Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister. This has been addressed with CAT models. IQtree provides a new family of C models that also address site heterogeneity. Given the extensive computational cost and concerns about overparameterization of CAT models [@whelan2016let], we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results. This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses. For computational efficiency, we first only applied Poisson+CAT models in every matrix with Choanoflagellata as outgroup. It should be noted that most Phylobayes runs were converged, although several large matrices have not reached convergence after at least a month's computational time.

Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in Choanimalia matrices. Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d matrix. Consistent with previous study, we recovered strong support of Porifera-sister in matrices of Philippe2008, Ryan2013, Nosenko2013 and Whelan2015. Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices and a weaker support of subsampled matrix of Simion2017. Ctenophora-sister were strongly supported in all other matrices. For the matrices have conflicted results between Poisson-CAT and C models, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model. We found strong support of Porifera-sister in holozoa matrices and Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups. Thus, our results show that both model choice and outgroup sampling play a major role in incongruent results of the root position of animal phylogeny in several key data matrices.

@@

The distribution of phylogenetic signal over the root position of animal phylogeny has never been quantified with site-heterogeneous models. To further explore the effect of model selection and outgroup choices in the root position of animal phylogeny, we next quantify the support of phylogenetic signal over two alternative hypotheses (T1: Ctenophora-sister (Fig. 1A); T2: Porifera-sister (Fig. 1B)) to three key data matrices with different outgroup choice and  models in both ML and BI framework (Fig XXPhylogenetic signal). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene (\Delta$GLS) or site (\Delta$SLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQtree or PhyloBayes). Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in site-homogeneous models. This finding is largely consistent with the past study that majority of phylogenetic signal is strongly favored in Ctenophora-sister hypothesis with site-homogeneous models in other data matrices [@shen2017contentious].

In contrast, the distribution of phylogenetic signal has never been quantified in site-heterogeneous models and compared with site-homogeneous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in two matrices from Ryan_2013 that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig. phylogenetic_signal B). It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C models lack of phylogenetic signal compared to site-homogeneous models in the node of root position of animal phylogeny. In contrast to the C models, the phylogenetic signal largely increased in Poisson-CAT model and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal. We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models. The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister increases in Holozoa and Choanimalia matrices.

We also examined the level of congruence of genes that are supported in each hypothesis across different analyses in Whelan2017 matrices. Consistent with the summary of \Delta$GLS analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice). In other words. our results strongly indicated that the distribution of phylogenetic signal of the root of animal phylogeny is largely sensitive to model selection (site-homogeneous vs. site-heterogeneous models). More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C models in these key data matrices. Thus, it is clear that recovering Porifera-sister requires more constrains, including more stringent model selection, outgroup choice, and even with gene sampling.

```{r phylogenetic_signal, fig.width=8, fig.height=15}
#au_tests %>%
#  unite(Matrix, c(manuscript, matrix, model)) %>%
#  group_by(Matrix, result) %>%
#  count() %>%
#  group_by(Matrix) %>%
#  mutate(percent = prop.table(n)) %>%

@@

au_tests = read_tsv("../trees_new/AU_test/summary.tsv")
